---
title: "Proteker2016"
author: "Yi-Ming Gan"
date: "`r Sys.Date()`"
knit: (function(inputFile, encoding) {
  rmarkdown::render(inputFile, encoding = encoding, output_dir = "html") })
output:
  html_document:
    df_print: paged
---

Transform Proteker 2016 data into sampling event core, occurrence extension and measurement or fact extension. The dataset is modeled this way because there is sediment sample that could not go into occurrence core.

## Read data 

### Event sheet

Event sheet was created by using unique records of [sampling event fields](https://rs.gbif.org/core/dwc_event_2022-02-02.xml) based on the sampling date.


```{r message=FALSE, warning=FALSE}
library(tidyverse)
# disable scientific notation so that coordinatePrecision of value 0.0001 will not become 1e-4
options(scipen = 999)

# read event from url
event <- read.csv("https://docs.google.com/spreadsheets/d/e/2PACX-1vTgsWJwmdxwhBtqo2JOAHMjQ0heA9jRtSB0ilfczTuVd80VBD79M3VYZPda69W9ALF286qd7BauJYqP/pub?gid=584519375&single=true&output=tsv", header = TRUE, sep = "\t", allowEscapes = FALSE, quote = "", check.names = FALSE)  
head(event)
names(event)

```

### Occurrence sheet

```{r}
# read occurrence from url
occ <- read.csv("https://docs.google.com/spreadsheets/d/e/2PACX-1vTgsWJwmdxwhBtqo2JOAHMjQ0heA9jRtSB0ilfczTuVd80VBD79M3VYZPda69W9ALF286qd7BauJYqP/pub?gid=0&single=true&output=tsv", header = TRUE, sep = "\t", allowEscapes = FALSE, quote = "", check.names = FALSE)  
head(occ)
names(occ)
```

## Create occurrence data frame

Data with fields of [occurrence core](https://rs.gbif.org/core/dwc_occurrence_2022-02-02.xml) is subset from the full dataset. This data frame will be written into `occurrence.txt` file.

```{r}
# exclude fields belong to sampling event core and measurement or fact extension
occ_df <- select(occ, -c(verbatimOccurenceID, measurementRemarks, year, month, day, coordinateUncertaintyInMeters, decimalLatitude, decimalLongitude, geodeticDatum, minimumDepthInMeters, maximumDepthInMeters, d13C, d15N, d34S), -contains("%"), -contains("C/N")) 
                 
head(occ_df)
names(occ_df)
```


## Read mof sheet 

MOF sheet contains the definition of all measurement types, BODC vocabularies and corresponding units. 

```{r}
# read measurementTypes sheet which has definition and unit for each measurmeent header
mof_fields <- read.csv("https://docs.google.com/spreadsheets/d/e/2PACX-1vTgsWJwmdxwhBtqo2JOAHMjQ0heA9jRtSB0ilfczTuVd80VBD79M3VYZPda69W9ALF286qd7BauJYqP/pub?gid=1897909656&single=true&output=tsv", header = TRUE, sep = "\t", check.names = FALSE)
mof_fields
```

## Simple function to subset a measurementType from wide table into long measurement or fact table

```{r message=FALSE}
#' Function to subset a measurementType from wide event table to long measurement or fact table
#'
#' @param data The data frame which contains measurement data in wide format.
#' @param mof_type The value should be "event" or "occurrence". If "event", mof variables will be extracted from event table and occurrenceID will not be assigned to the MOF data frame. If "occurrence", mof variables and occurrenceID be extracted from occurrence table and occurrenceID will be used to create the eMOF record.
#' @param mof_fields The data frame which contains measurement columnName of `data`, measurementType and measurementUnit
#' @param measurement_col A string which is the field name of data frame that contains the measurement value. Will be used to populate the measurementValue field.
#
create_emof <- function(data = data, mof_type, mof_fields = mof_fields, measurement_col) {
  # create a measurement or fact table of a single measurement type from the data
  # subset from full dataset
  if (mof_type =="event") {
    mof <- data %>% 
      # remove rows of measurement which are NULL
      filter(!is.na(!!as.symbol(measurement_col))) %>%  # !!as.symbol is required as column name is passed as string
      select(eventID, !!as.symbol(measurement_col), measurementRemarks) %>%  # occurrenceID is absent because this is from an event table
      mutate(occurrenceID="")
  } else if (mof_type =="occurrence") {
    mof <- data %>% 
    # remove rows of measurement which are NULL
    filter(!is.na(!!as.symbol(measurement_col))) %>%  # !!as.symbol is required as column name is passed as string
    select(eventID, occurrenceID, !!as.symbol(measurement_col), measurementRemarks) # select occurrenceID here
  } else { break }
  
  # pull measurementType, measurementTypeID, measuremntUnit and measurmentUnitID from mof sheet based on the column name
  measurement_type <- mof_fields %>% filter(columnName == measurement_col) %>% pull(measurementType)
  measurement_type_id <- mof_fields %>% filter(columnName == measurement_col) %>% pull(measurementTypeID)
  measurement_unit <- mof_fields %>% filter(columnName == measurement_col) %>% pull(measurementUnit)
  measurement_unit_id <- mof_fields %>% filter(columnName == measurement_col) %>% pull(measurementUnitID)

  # create mof table (long table)
  mof_df <- data.frame(
    eventID = mof[, "eventID"],
    occurrenceID = mof[, "occurrenceID"],
    measurementType = measurement_type,
    measurementTypeID = measurement_type_id,
    measurementValue = mof[, measurement_col],
    measurementUnit = measurement_unit,
    measurementUnitID = measurement_unit_id,
    measurementRemarks = mof[, "measurementRemarks"]
  )  
  return(mof_df)
}

```


## Subsets of different measurements

Create subsets of different measurements

```{r}
# event MOF
event_water_temp <- create_emof(data = event, mof_type = "event", mof_fields = mof_fields, measurement_col = "water temperature (°C)")
event_d13C <- create_emof(data = event, mof_type = "event", mof_fields = mof_fields, measurement_col = "d13C")
event_d15N <- create_emof(data = event, mof_type = "event", mof_fields = mof_fields, measurement_col = "d15N")
event_d34S <- create_emof(data = event, mof_type = "event", mof_fields = mof_fields, measurement_col = "d34S")
event_percent_C <- create_emof(data = event, mof_type = "event", mof_fields = mof_fields, measurement_col = "%C")
event_percent_N <- create_emof(data = event, mof_type = "event", mof_fields = mof_fields, measurement_col = "%N")
event_percent_S <- create_emof(data = event, mof_type = "event", mof_fields = mof_fields, measurement_col = "%S")
event_C_N_ratio <- create_emof(data = event, mof_type = "event", mof_fields = mof_fields, measurement_col = "C/N")

# occurrence MOF
occ_d13C <- create_emof(data = occ, mof_type = "occurrence", mof_fields = mof_fields, measurement_col = "d13C")
occ_d15N <- create_emof(data = occ, mof_type = "occurrence", mof_fields = mof_fields, measurement_col = "d15N")
occ_d34S <- create_emof(data = occ, mof_type = "occurrence", mof_fields = mof_fields, measurement_col = "d34S")
occ_percent_C <- create_emof(data = occ, mof_type = "occurrence", mof_fields = mof_fields, measurement_col = "%C")
occ_percent_N <- create_emof(data = occ, mof_type = "occurrence", mof_fields = mof_fields, measurement_col = "%N")
occ_percent_S <- create_emof(data = occ, mof_type = "occurrence", mof_fields = mof_fields, measurement_col = "%S")
occ_C_N_ratio <- create_emof(data = occ, mof_type = "occurrence", mof_fields = mof_fields, measurement_col = "C/N")

# example
head(event_d13C)
head(occ_d13C)

```

Bind all the subsets into one data frame to form the measurement or fact table.

```{r}
mof_df <- rbind(
  event_water_temp,
  event_d13C,
  event_d15N, 
  event_d34S, 
  event_percent_C, 
  event_percent_N, 
  event_percent_S, 
  event_C_N_ratio, 
  occ_d13C, 
  occ_d15N, 
  occ_d34S, 
  occ_percent_C, 
  occ_percent_N, 
  occ_percent_S, 
  occ_C_N_ratio
  
)

dim(mof_df)  # dimension of the mof long table
head(mof_df)  # preview mof table
```

## Create measurementID

Add a column of unique measurementID per record

```{r}
# add a sequential number to eventID for this
mof_df <- mof_df %>% mutate(measurementID=paste(eventID, 1:nrow(mof_df), sep = "_"), .before=eventID)
head(mof_df)
```



## Write data frames to files

```{r message=FALSE}
library(here)
library(readr)
require(dplyr)    

outfile_dir <- here("data/processed")
event_file <- file.path(outfile_dir, "event.txt")
occ_file <- file.path(outfile_dir, "occurrence.txt")
mof_file <- file.path(outfile_dir, "measurementOrFact.txt")

write_tsv(event, event_file, na = "", quote = "none", escape = "none")
write_tsv(occ_df, occ_file, na = "", quote = "none", escape = "none")
write_tsv(mof_df, mof_file, na = "", quote = "none", escape = "none")
```
